---
title: "Predicting wins and methods of victory for UFC fighters using tidymodels"
output: html_document
---

The objective is to determine which stats are most predictive of a UFC fighter winning a fight and to train a model to predict the method of victory using the stats of a fighter.

The main method of analysis is the [tidymodels](https://www.tidymodels.org/) package. The UFC data was found on [Kaggle](https://www.kaggle.com/datasets/remypereira/mma-dataset-2023-ufc?resource=download) (see the ReadMe). The full data set contains four files related via ID columns for the event, fight, fighter, and fight stat.

```{r load, message=FALSE, warning=FALSE}
library(tidymodels)
library(tools)
library(vip)

# List our data files
files <- list.files(".data", full.names = TRUE)

# Add each of the data files to our Global environment
walk(files, function(f) {
  nm <- file_path_sans_ext(basename(f))
  df <- read.csv(f)
  assign(nm, df, envir = .GlobalEnv)
})

# Print filenames and names of ID columns
cols <- vector()
for (obj in ls()) {
  col <- NULL
  if (grepl("_data$", obj)) {col <- grep("_id", colnames(get(obj)), value = TRUE)}
  if(!is.null(col)) {cols <- c(cols, col)}
}

cat("Files: ", paste(basename(files), collapse = ", "), "\n")
cat("ID columns: ", paste(unique(cols), collapse = ", "), "\n")
```

First we will use the ID columns to join our files into a single dataframe, and we will select the stats we want to use to predict the fight result. The chosen stats describe fighter size, stance, strikes thrown and landed, takedowns attempted and achieved, knockdowns, submission attempts, reversals, and control time.

```{r}
# Select and merge the relevant data for our models
data <- left_join(ufc_event_data, ufc_fight_data, by = "event_id")
data <- left_join(data, ufc_fight_stat_data, by = "fight_id")
data <- left_join(data, ufc_fighter_data, by = "fighter_id")

model_data <- select(
  data, 
  fight_id, 
  fighter_id, 
  fight_stat_id, 
  fighter_id, 
  fighter_height_cm, 
  fighter_weight_lbs, 
  fighter_reach_cm, 
  fighter_stance,
  total_strikes_att, 
  total_strikes_succ, 
  sig_strikes_att, 
  sig_strikes_succ, 
  knockdowns, 
  takedown_att, 
  takedown_succ, 
  submission_att, 
  reversals, 
  ctrl_time, 
  result,
  winner
) 

# Reformat control time from format "MM:SS" to total seconds
model_data <- model_data %>% 
  mutate(
    ctrl_time = case_when(
      ctrl_time == "--" ~ NA_character_,
      TRUE ~ ctrl_time),
    ctrl_time_seconds = case_when(
      is.na(ctrl_time) ~ NA_real_,
      TRUE ~ as.numeric(substr(ctrl_time, 1, regexpr(":", ctrl_time)-1)) * 60 + as.numeric(substr(ctrl_time, regexpr(":", ctrl_time)+1, nchar(ctrl_time))))) %>%
  select(-ctrl_time)

# Clean and preprocess data
model_data <- na.omit(model_data)
model_data <- mutate(model_data, winner = ifelse(winner == fighter_id, 1, 0))
model_data <- relocate(model_data, result, winner, .after = last_col())
model_data <- arrange(model_data, fight_id, fight_stat_id)
```

Let's determine which stats are most predictive of a fighter winning their fight. First we have to create the model. We will do this using [tidymodels](https://www.tidymodels.org/) functions split our train/test data, create our recipe and pre-processing steps, set our model engine and specifications, create our workflow and finally fit the model. We use the vip() function from the [vip](https://github.com/koalaverse/vip/) package to asses the importance of each stat in predicting the fight result.

```{r}
set.seed(123)
tidymodels_prefer()

# Create training/testing split
split <- initial_split(model_data, prop = 0.8, strata = winner)
train <- training(split)
test <- testing(split)

# Create recipe
recipe <- recipe(
  winner ~ fighter_height_cm + fighter_weight_lbs + fighter_reach_cm + fighter_stance + 
           total_strikes_att + total_strikes_succ + sig_strikes_att + sig_strikes_succ + 
           knockdowns + takedown_att + takedown_succ + submission_att + reversals + ctrl_time_seconds, 
  data = train) %>%
  step_mutate(winner = factor(winner, levels = c("0", "1"))) %>%
  step_mutate(fighter_stance = na_if(fighter_stance, "")) %>%
  step_unknown(fighter_stance) %>%
  step_dummy(fighter_stance) %>%
  step_normalize(all_numeric_predictors())

# Create logistic regression model
spec <- logistic_reg(penalty = 0.01) %>%
  set_engine("glmnet") %>%
  set_mode("classification")

# Create workflow
workflow <- workflow() %>%
  add_recipe(recipe) %>%
  add_model(spec)

# Fit model
fit <- workflow %>% fit(data = train)

# Calculate variable importance
importance <- fit %>%
  extract_fit_parsnip() %>%
  vip()

print(importance)
```

Evidently (and maybe not surprisingly) the most important statistics that predict whether a fighter will win or lose a fight are significant strikes, both attempted and successfully landed, and how many of those strikes resulted in a knockdown on their opponent. Total control time, takedowns attempted and achieved, and submission attempts also show high importance.

We see from our data that there are only five ways to win a fight.

```{r}
cat("Win by: ", paste(unique(model_data$result), collapse = ", "), "\n")
```

This raises the next question: Can we use a fighter's stats to predict by which method they win each of their fights?

The possible fight outcomes and their relative proprtions in the data are shown below.

```{r}
model_data %>%
  count(result) %>%
  mutate(proportion = n/sum(n))
```

Let's again use [tidymodels](https://www.tidymodels.org/) to try to predict the fight outcome from the given stats.

```{r}
# Filter for winning fights only and select relevant features
model_data <- model_data %>%
  filter(winner == 1) %>%
  select(-fight_id, -fighter_id, -fight_stat_id, -winner) %>%
  mutate(result = as.factor(result))

# Split the data
set.seed(123)
split <- initial_split(model_data, prop = 0.8, strata = result)
train <- training(split)
test <- testing(split)

# Create cross-validation folds
folds <- vfold_cv(train, v = 5, strata = result)

# Create a recipe for preprocessing
recipe <- recipe(result ~ ., data = train) %>%
  step_string2factor(fighter_stance) %>%
  step_dummy(fighter_stance) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_zv(all_predictors())

# Create a random forest model specification
rf_spec <- rand_forest(
  trees = 1000,
  min_n = tune(),
  mtry = tune()
) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("classification")

# Create a workflow
rf_workflow <- workflow() %>%
  add_recipe(recipe) %>%
  add_model(rf_spec)

# Define the parameter grid
rf_grid <- grid_regular(
  min_n(range = c(2, 10)),
  mtry(range = c(2, 8)),
  levels = 5
)

# Tune the model
rf_tune_results <- tune_grid(
  rf_workflow,
  resamples = folds,
  grid = rf_grid,
  metrics = metric_set(accuracy, roc_auc)
)

# Get best parameters
best_params <- select_best(rf_tune_results, metric = "accuracy")

# Finalize the workflow with best parameters
final_workflow <- rf_workflow %>%
  finalize_workflow(best_params)

# Fit the final model
final_fit <- final_workflow %>%
  fit(data = train)

# Make predictions on test set
predictions <- predict(final_fit, test)
predictions_prob <- predict(final_fit, test, type = "prob")

# Combine actual and predicted values
results <- bind_cols(
  test %>% select(result),
  predictions,
  predictions_prob
)

# Calculate performance metrics
conf_mat <- conf_mat(results, truth = result, estimate = .pred_class)
accuracy <- accuracy(results, truth = result, estimate = .pred_class)

# Get variable importance
final_fit %>%
  extract_fit_parsnip() %>%
  vip()

# Print results
print("Confusion Matrix:")
print(conf_mat)
print("Accuracy:")
print(accuracy)
```

Now that we know which stats are highly predictive of the pathway to victory, we will filter for the fights where the winner had high values for each stat and see if excelling in a particular stat results in a higher probability of a certain fight outcome.

```{r}
# Function to analyze fights where winner excelled in a particular stat
analyze_high_values <- function(data, variable) {
  threshold <- quantile(data[[variable]], 0.75, na.rm = TRUE)
  
  high_value_results <- data %>%
    filter(!!sym(variable) >= threshold) %>%
    count(result) %>%
    mutate(proportion = n / sum(n)) %>%
    arrange(desc(proportion))
  
  top_result <- high_value_results %>%
    slice(1)
  
  return(data.frame(
    variable = variable,
    threshold = threshold,
    most_common_result = top_result$result,
    probability = top_result$proportion
  ))
}

# Top 10 important variables
important_vars <- c("submission_att", "total_strikes_att", "total_strikes_succ", "sig_strikes_att", "knockdowns", "ctrl_time_seconds", "sig_strikes_succ", "takedown_att", "takedown_succ", "fighter_reach_cm")

# Analyze each variable
results_analysis <- map_df(important_vars, ~analyze_high_values(model_data, .x))

# Print results in a nice format
results_analysis %>%
  mutate(
    probability = round(probability * 100, 1),
    threshold = round(threshold, 2)) %>%
  arrange(desc(probability)) %>%
  mutate(interpretation = paste0(
    "When ", variable, " is above ", threshold, 
    ", the most likely result is ", most_common_result,
    " (", probability, "% of cases)")) %>%
  pull(interpretation)
```

Evidently (and again, maybe not surprisingly), having an impressive stat for any kind of striking, takedown, or control time suggests that the fighter will win by decision. Two exceptions to this are knockdowns, which suggest a fighter will win by KO/TKO, and submission attempts, which suggest a fighter will win by submission. A larger reach also increases the probability of winning by KO/TKO.